Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-05-24 joint_inf.tex
Moving appendix to very end.
pass through

Classical maximum-likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
Recently \cite{Neher2017} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
This latter step is simpler and more computationally efficient when conditioning on ancestral sequences compared to the case in which ancestral sequences are marginalized out.

However, existing consistency proofs for phylogenetics \cite{RoyChoudhury2015-ta} do not apply to the setting of joint inference of trees and ancestral sequences.
There are examples in statistics, such as the Neyman-Scott paradox \cite{Neyman1948-tt}, in which estimation of additional parameters destroys consistency.
This raises the question of whether joint inference of trees and ancestral sequences enjoys good statistical properties.

In this paper we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we follow a strategy of using simple models and four taxa trees.
Here we use a binary symmetric model and simulate on the ``Farris zone'' \cite{Siddall1998-hq} tree.
We use bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference converges on the wrong tree.

@@

We will be considering the standard phylogenetic likelihood on unrooted phylogenetic trees (cite Felsenstein book).
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ distributed as $\ancestralStateColumnRV$.

@@

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
%das: I forgot that standard likelihoods are written with the parameters on the left.
%Since H is like a hidden variable, I went for an expectation-maximization notation, grouping them with the observations.
%It seems a little wonky, though, when we start having maximum likelihood ancestral states depending on \tau,t.
%Oh well...
L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \argmax_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\tau, t; \fullAlignment).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
We can perform this by computing the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates),
\end{equation}
then estimating the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment)
\end{equation}
while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, a given $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
Indeed, we can start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.

Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the power set.
For topology $\tau$, we also define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
The probability of observing an ancestral state split, conditional on a site split, will not be invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
The number of unique ancestral state splits is $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$.

With this split formulation, define mappings from patterns to splits as described above:
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet).
Given a site-pattern-valued random variable $\alignmentColumnRV$ and an ancestral-state-valued random variable $\ancestralStateColumnRV$, define random variables
\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV).
% Need to introduce \siteSplit_j and \ancestralSplit_k , no?
We can now write the $i$th factor of \eqref{eq:full_likelihood} as
    P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
for some $j$ and $k$.
Taking the maximum across internal states yields
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
This allows us to write the likelihood in \eqref{eq:profile_likelihood} as a product over site patterns as opposed to sites, i.e.,
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
%EM Indexing broken here: we are taking a product over i, but i doesn't appear in the expression.
    %das: This is intentional, though: it allows us to proceed to the following line, grouping factors by site pattern since their probabilities don't depend on $i$. I can try to think of how to make this more clear, but I thought it was a pretty key step to be able to pass to site patterns, and the reason for the machinery involved in defining ancestral state partitions.
%EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} P(\ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
%EM think we should say "where the site split or its complement appears."?
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

@@

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments can be described as
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips at which a character is assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Had we not identified character complements, we would need to consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
but the symmetry of the binary character model allows us to focus just on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
We cannot similarly identify complements in this case since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
%EM Would it be worth enumerating the site splits ahead of time? It seems a little funky to be calling them j each time.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
Thus the set of most likely ancestral states contains a single element in this case, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_6=\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2.
$$
In this case, the set of most likely ancestral states is $\ancestralSplitPartition_6(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_6\in\ancestralSplitPartition_6(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_6 \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3.
$$

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
%EM clarify that we aren't assuming that the x and y in t are the same as those appearing in t^*?
Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.

>>>>

Classical maximum-likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
Recently, \cite{Neher2017} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
This is attractive from a computational perspective: such joint ML inference proceeds according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
This latter conditional optimization is simpler and more computationally efficient than marginalizing out ancestral sequences and then performing optimization.

Existing consistency proofs for phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-ta} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we follow a strategy of using simple models and four taxa trees.
We use a binary symmetric model and simulate on the ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference converges to the wrong tree.

@@

We consider the standard phylogenetic likelihood on unrooted phylogenetic trees \cite{Felsenstein2004}.
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
Assume we observe $\nCols$ samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.

@@

For a topology $\tau$ and branch lengths $t$, we make the usual assumption of independence between sites, obtaining the likelihood
\begin{equation}
\label{eq:full_likelihood}
%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
%das: I forgot that standard likelihoods are written with the parameters on the left.
%Since H is like a hidden variable, I went for an expectation-maximization notation, grouping them with the observations.
%It seems a little wonky, though, when we start having maximum likelihood ancestral states depending on \tau,t.
%Oh well...
L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \argmax_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\tau, t; \fullAlignment).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
We compute the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates),
\end{equation}
then estimate the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment),
\end{equation}
using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and \eqref{eq:profile_likelihood_topology_bl} recovers the joint ML values of the unknown parameters.
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
Indeed, we start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.

Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the power set.
For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
The probability of observing an ancestral state split, conditional on a site split, is not invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
The number of unique ancestral state splits is $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$.

With this split formulation, define mappings from patterns to splits as
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet).
$$
Given a site pattern--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, define random variables
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV)
$$
that take corresponding realizations $\siteSplit_j$ and $\ancestralSplit_k$ for some $j$ and $k$.
We write the $i$th factor of \eqref{eq:full_likelihood} as
\begin{align*}
    P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t).
\end{align*}
Taking the maximum across internal states yields
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} P(\ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

@@

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips at which a character is assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
We cannot similarly identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_6=\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2.
$$
Here, the set of most likely ancestral states is $\ancestralSplitPartition_6(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_6\in\ancestralSplitPartition_6(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_6 \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3.
$$

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ constrains the bottom three branches to share the same parameter, the classical construction of this topology.
